---
title: "Reshaping Data"
subtitle: "Using `tidyr` and `data.table`"
institute: "QMER Workshop"
date: "Fall 2022 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: "qmer-xaringan.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
---


```{r setup, include=FALSE}
library(knitr)
options(htmltools.dir.version = FALSE)
opts_chunk$set(echo = FALSE, comment = NULL)
```
class: large
## QMER: What we do

QMER is a open learning community serving all of Auburn University by providing:

* Support for Statistics, Project Management, and Workflow

* Professional skills not taught in your courses

* Research labs, teaching opportunities

* Large scale public data sets

---
class: large
## Outline of the Workshop

* What is reshaping data and why do we need to do it?
* Brief overview of ways to reshape data in R
* Demonstration of wide to long with `tidyr`
* Demonstration of wide to long with `data.table`
* Real-life example with `data.table`

---
class: inverse middle center large

# The What and Why of Reshaping Data

---
class: large
## What is Reshaping Data

* Reshaping data entails re-organizing data without changing changing the information represented.
* Contrast this with manipulating data, which entails altering information or creating new variables.
* Or contrast this with merging, which also adds information.

---
### Toy Example of Wide and Long Data
```{r, echo=FALSE}
library(tidyr)
data.wide <- data.frame(id = 1:4, 
                        math_k = c(25, 27, 21, 30),
                        math_1st = c(31, 32, 28, 38),
                        math_2nd = c(36, 34, 32, 48))

data.long <- pivot_longer(data.wide, 
                          cols = contains("math"),
                          names_to = "grade",
                          names_prefix = "math_",
                          values_to = "math")
data.long$grade <- factor(data.long$grade)
```

.pull-left[

```{r}
kable(data.wide, caption = "wide data", )
```
]

.pull-right[


```{r}
kable(data.long, caption = "long data")
```

]

---
class:  large

## Ways to Reshape Data in R

* `reshape()` function in **base R**
* `melt()` and `cast()` function in **reshape2** package
* **tidyr** package from the tidyverse
* **data.table** package has a faster version of `melt()`


---
class: large

## Reshaping with **tidyr**

[tidyr package documentation](https://tidyr.tidyverse.org/index.html)

[tidyr pivoting tutorial](https://tidyr.tidyverse.org/articles/pivot.html)

---
class: large

## Reshaping with `data.table`

[data.table package documentation](https://rdatatable.gitlab.io/data.table/index.html)

[data.table reshaping vignette](https://rdatatable.gitlab.io/data.table/index.html)